Class differences in social networks: Evidence from a referral experiment

### 2025-07-29 writing/manuscript.tex
v1.1.1

Our study takes place at UNAB, a medium-sized private university in Bucaramanga, Colombia with approximately 6,000 enrolled undergraduate students. The university's student body is remarkably diverse, with about 35\% of students classified as low-SES and 15\% as high-SES. This diversity at UNAB provides a unique research setting. Colombian society is highly unequal and generally characterized by limited interaction between social classes, with different socioeconomic groups separated by education and geographic residence.\footnote{Colombia has consistently ranked as one of the most unequal countries in Latin America \citep{barretoherrera2024regional}, with the richest decile earning 50 times more than the poorest decile \citep{comisioneconomicaparaamericalatinayel2023}. This economic disparity is reflected by a highly stratified society with significant class inequalities and limited class mobility \citep{angulo2012movilidad,garcia2015loteria}.} Despite significant financial barriers, many lower and middle-SES families prioritize university education for their children \citep[p. 103]{hudson_colombia_2010}. UNAB represents one of the few environments in Colombia where sustained inter-SES contact occurs naturally (see Figure \ref{school_quality_income}).

@@

To operationalize SES in our study, we rely on Colombia's established classification system. In 1994, Colombia introduced a nationwide system that divides the population into 6 strata based on housing characteristics and neighborhood amenities.\footnote{Initially designed for utility subsidies from higher strata (5 and 6) to support lower strata (1 to 3), it now extends to university fees and social program eligibility. Stratum 4 neither receives subsidies nor pays extra taxes. This stratification system largely aligns with and potentially reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}.} We use this classification as our measure of SES in the experiment. Students in strata 1 to 2 are categorized as low-SES, strata 3 to 4 as middle-SES, and those in strata 5 to 6 as high-SES.

For our experiment, we invited via email all 4,417 UNAB undergraduate students who had completed their first year at the university at the time of recruitment. The 837 students who joined (19\%) vary in terms of their academic programs, SES, and progress in their studies. This experimental setup provides a unique opportunity for collaborative inter-class contact on equal status, whose positive effects on reducing discrimination are causally documented \citep{rao2019familiarity, mousa2020building,lowe2021types}.

The institutional setup at UNAB sets up the social contact structure and facilitates the network analysis. Undergraduate programs at UNAB are spread across two semesters, with each individual course lasting one semester. Students take between 5 to 7 courses per semester, with programs lasting anywhere between 4 to 12 semesters (2 to 6 years). Medicine, the largest program by size at UNAB, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters.

\section{Design} \label{s3}

We designed an experiment to assess peer referral behavior from an SES perspective and to causally evaluate the effect of different incentive structures on referrals. The study design consists of a single online experiment organized at the university level (see Figure \ref{timeline}). The instructions are provided in Appendix \ref{instructions}.

\begin{figure}[H]
\centering
\caption{Experiment Timeline}\label{timeline}
\includegraphics[width=\linewidth]{../figures/timeline.jpg}
\begin{tablenotes}
\footnotesize
\item[] \textit{Note:} Participants first report beliefs about their own national university entry exam performance, then recommend peers for each academic area. In the final part, they report beliefs about their recommendations' performance and provide demographic information. This order is implemented for all participants.
\end{tablenotes}
\end{figure}

\subsection{Productivity measures}
To establish an objective basis for referral productivity, we use national university entry exam scores (SABER 11). These scores provide pre-existing, comparable measures of ability across two domains relevant for the labor market. By using existing administrative data, we eliminate the need for additional testing and ensure that all eligible students have comparable productivity measures. The scores we use in this experiment consist of critical reading and mathematics parts.

Critical reading evaluates competencies necessary to understand, interpret, and evaluate texts found in everyday life and broad academic fields (e.g., history). This measures students' ability to comprehend and critically evaluate written material. Mathematics assesses students' competency in using undergraduate level mathematical tools (e.g., reasoning in proportions, financial literacy). This captures quantitative reasoning and problem-solving abilities.

For each area, we calculate percentile rankings based on the distribution of scores among all currently enrolled UNAB students, providing a standardized measure of relative performance within the university population.

\subsection{Referral task}
Building on these productivity measures, our main experimental task involves peer recommendations. After eliciting beliefs about their own performance, participants engage in incentivized peer recommendations. For both test areas (critical reading and mathematics), participants recommend one peer they believe excels in that domain. We first present an example question from the relevant test area to clarify what skills are being assessed. Participants then type the name of their recommended peer, with the system providing autocomplete suggestions from enrolled students who have taken the test (see Figure \ref{interface}).

@@

Participants can only recommend students with whom they have taken at least one class during their university studies. This requirement ensures that referrals are based on actual peer interactions and overlaps with the enrollment network that we construct. The order in which participants make recommendations across the two areas is randomized.

We incentivize referrals using a productivity-based payment scheme that rewards participants for recommending highly ranked peers. Referrers earn increasing monetary rewards as the percentile ranking of their recommendation increases (see Figure \ref{piece_rate_by_score}). This payment structure provides strong incentives to screen for highly ranked peers, with potential earnings up to \$60 per recommendation. We multiply the piece rate coefficient associated with the percentile rank by the actual test scores of the recommendation to calculate earnings.\footnote{Due to the selection into the university, the actual test score distribution has limited variance. Below a certain threshold students cannot qualify for the institution and choose a lower ranked university, and above a certain threshold they have better options to choose from.}

@@

\subsection{Treatment variation}
To examine how different incentive structures affect referral behavior, we implement a treatment that varies payment to recommended peers. We implement a between-subjects treatment that varies whether the recommended peer also receives payment. In the \textbf{Baseline} treatment, only the referrer can earn money based on their recommendation's productivity. The \textbf{Bonus} treatment adds an additional fixed payment of \$25 to any peer who is recommended in the randomly selected area for payment. This payment is independent of the peer's actual productivity (see Figure \ref{tab:treatments}).

\begin{table}[htbp]
  \centering
  \begin{threeparttable}
  \caption{Incentive structure by treatment}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c @{}}
  \toprule
  & \multicolumn{1}{c}{\textbf{Baseline}} & \multicolumn{1}{c}{\textbf{Bonus}} \\
  \midrule
  Referrer (sender) & Productivity-based & Productivity-based \\
  Recommendation (receiver) & No payment  & Fixed reward  \\
  \bottomrule
  \end{tabular*}
  \label{tab:treatments}
  \end{threeparttable}
\end{table}

Participants are informed about their treatment condition before making recommendations through both video and text instructions. The treatment is assigned at the individual level, allowing us to compare referral outcomes across conditions.

\subsection{Belief elicitation}
To understand participants' information and expectations, we collect incentivized beliefs at key points. We elicit incentivized beliefs at two points in the experiment. First, before making referrals, participants report their beliefs about their own percentile ranking in each test area. Second, after making each referral, participants report their beliefs about their recommendation's percentile ranking. For both belief elicitation tasks, participants earn \$5 if their guess is within 7 percentiles of the true value. This tolerance level is expected to balance precision with the difficulty of the task.

@@

\section{Sample, Incentives, and Procedure} \label{s4}
We invited all 4,417 UNAB undergraduate students who had completed their first year at the university at the time of recruitment to participate in our experiment. A total of 837 students participated in the data collection, yielding a 19\% response rate. Our final sample consists of 734 individuals who referred peers with whom they had taken at least one class together, resulting in an 88\% success rate for the sample. We randomly allocated participants to either \textbf{Baseline} or \textbf{Bonus} treatments.

@@

The experiment was conducted online through Qualtrics, with participants recruited from active UNAB students. To ensure data quality while managing costs, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion, plus potential earnings from one randomly selected belief question (up to \$5) and one randomly selected recommendation question (up to \$60). This structure resulted in maximum total earnings of \$82. The average time to complete the survey was 30 minutes, with an average compensation of \$80 for the one in ten participants randomly selected for payment. Payment processing occurred through online banking platform Nequi within 15 business days of participation.

@@

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Selection into the experiment}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
  & \multicolumn{1}{c}{\textbf{University}} & \multicolumn{1}{c}{\textbf{Sample}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\
  \midrule
  Reading score & 62.651 & 65.183 & 0.000 \\
  Math score & 63.973 & 67.477 & 0.000 \\
  GPA & 3.958 & 4.012 & 0.000 \\
  Low-SES & 0.343 & 0.410 & 0.000 \\
  Middle-SES & 0.505 & 0.499 & 0.763 \\
  High-SES & 0.153 & 0.091 & 0.000 \\
  Female & 0.567 & 0.530 & 0.060 \\
  Age & 21.154 & 20.651 & 0.000 \\
  \midrule
  Observations & 4,417 & 734 &  \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table compares characteristics between the university and the experimental sample. $p$-values for binary outcomes (Low-SES, Med-SES, High-SES, Female) are from two-sample tests of proportions; for continuous variables, from two-sample $t$-tests with unequal variances. All reported \textit{p}-values are two-tailed.
  \end{tablenotes}
  \label{tab:selection}
  \end{threeparttable}
\end{table}

>>>>

Our experiment took place in Colombia, a country that consistently ranks highly in terms of economic inequality. The richest decile of Colombians earn 50 times more than the poorest decile \citep{barretoherrera2024regional,comisioneconomicaparaamericalatinayel2023}. This economic disparity creates profound differences in outcomes across SES groups in terms of education, geographic residence, language, manners, and social networks \citep{garciavillegas2021dimension, angulo2012movilidad,garcia2015loteria}. While these patterns are not atypical and exist elsewhere, Colombia's pronounced inequality makes economic, educational, and cultural differences across SES particularly visible and thus provides an ideal setting to study SES biases in referral selection.

We rely on Colombia's established estrato classification system to measure SES in our study. In 1994, Colombia introduced a nationwide system that divides the population into six strata based on  ``similar social and economic characteristics'' \citep[p. 102]{hudson_colombia_2010}. Designed for utility subsidies from higher strata to support lower strata, the system aligns with and reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}. It is widely used by policymakers and in official statistics \citep{fergusson2021desigualdad}. Using the estrato system, we categorize students in strata 1-2 as low-SES, strata 3-4 as middle-SES, and strata 5-6 as high-SES.

Colombia's educational segregation typically prevents meaningful interaction between socioeconomic groups, as wealthy families attend exclusive private schools while poorer families access lower-quality public or ``non-elite'' private institutions (see Figure \ref{school_quality_income}). Our study takes place in a non-elite private university which attracts students across the socioeconomic spectrum: The university's student body comprises 35\% low-SES, 50\% middle-SES, and 15\% high-SES students.\footnote{Government statistics reveal less than 5\% of the population is high-SES \citep[p. 103]{hudson_colombia_2010}.} This diversity provides opportunities for different SES groups to meet and interact within the same institutional framework.

The partner university creates conditions for contact on equal status. All students pay the same fees based on their program choices, and less than 5\% of students receive scholarships. The student body is mostly urban and has comparable university entry exam scores due to the entrance exam \textcolor{red}{ADD NUMBERS}. These additional factors make our setting appropriate to study the effects of contact on intergroup discrimination.

% These factors typically facilitate reducing bias between interacting groups \citep{allport1954nature,rao2019familiarity,mousa2020building,lowe2021types}

% For our experiment, we invited via email all 4,417 undergraduate students who had completed their first year at the university at the time of recruitment. These students vary in terms of their academic programs \textcolor{red}{ADD NUMBERS}, SES \textcolor{red}{ADD NUMBERS}, and progress in their studies \textcolor{red}{ADD NUMBERS}.

% Undergraduate students take between 5 to 7 courses per semester, with programs lasting anywhere between 4 to 12 semesters (2 to 6 years). Medicine, the largest program by size at the university, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters.

% 837 students joined the experiment (19\%).

@@

\section{Design} \label{s3}
We designed an online experiment to assess peer referral selection from an SES perspective and to evaluate the causal effect of providing a bonus to referral candidates. The experimental design consisted of two incentivized tasks administered in the following sequence to all participants: First, participants completed belief elicitation tasks about their own performance on the national university entry exam. Second, they completed the main referral task, nominating peers based on exam performance in two academic areas. Finally, participants reported beliefs about their referrals' performance and provided demographic information. This structure allowed us to measure both the accuracy of participants' beliefs and their referral behavior under controlled incentive conditions. Figure \ref{timeline} shows the experimental timeline, and detailed instructions are provided in Appendix \ref{instructions}.

\begin{figure}[H]
\centering
\caption{Experimental Timeline}\label{timeline}
\includegraphics[width=\linewidth]{../figures/timeline.jpg}
\begin{tablenotes}
\footnotesize
\item[] \textit{Note:} Participants first reported beliefs about their own university entry exam performance, then made referrals for each academic area, and finally reported beliefs about their referrals' performance and provided demographics.
\end{tablenotes}
\end{figure}

\subsection{Performance measures}
To establish an objective basis for referral performance, we use national university entry exam scores (SABER 11). All Colombian high school students take the SABER 11 exam at the end of their final year as a requirement for university admission. These scores provide pre-existing, comparable measures of performance. By using existing administrative data, we also ensure that all eligible students have comparable performance measures.

The SABER 11 exam consists of five areas (critical reading, mathematics, natural sciences, social sciences, and English). We focus on critical reading and mathematics as these represent fundamental skills most relevant across academic programs and future employment. Critical reading evaluates competencies necessary to understand, interpret, and evaluate texts found in everyday life and broad academic fields (e.g., history), while mathematics assesses students' competency in using undergraduate level mathematical tools (e.g., reasoning in proportions, financial literacy). These together capture performance in comprehending and critically evaluating written material as well as reasoning and problem-solving abilities.

For each area, we calculate percentile rankings based on the distribution of scores among all currently enrolled students, providing a standardized measure of relative performance within the university population.

\subsection{Referral task}
The main experimental task involves making referrals among peers. For both exam areas (critical reading and mathematics), participants refer one peer they believe excels in that area. We provide an example question from the relevant exam area to clarify the skills that are being assessed. Participants type the name of their preferred candidate to make a referral. To avoid issues with recall, the interface provides autocomplete name and program suggestions from the administrative database (see Figure \ref{interface}).

@@

Participants can only refer students with whom they have taken at least one class during their university studies. This requirement ensures that referrals are based on actual peer interactions. We randomize the order in which participants make referrals across the two exam areas.

We incentivize referrals using a piece rate payment structure. Referrers earn increasing payments as the percentile ranking of their recommendation increases (see Figure \ref{piece_rate_by_score}). We multiply the piece rate coefficient associated with the percentile rank by the actual exam scores of the recommendation to calculate earnings. This payment structure provides strong incentives to refer highly ranked peers with potential earnings going up to \$60 per referral. \footnote{Note that due to the selection into the university, the actual exam score distribution has limited variance. Below a certain threshold students cannot qualify for the institution and choose a lower ranked university, and above a certain threshold they have better options to choose from.}

@@

\subsection{Bonus Treatment}
To examine how different incentive structures affect referral selection, we randomly assign a fixed bonus payment for students who get a referral. In the \textbf{Baseline} treatment, only the referrer can earn money based on their referral's performance. The \textbf{Bonus} treatment adds an additional fixed payment of \$25 to the peer who gets the referral. This payment is independent of the referred candidate's actual performance (see Table \ref{tab:treatments}).

\begin{table}[htbp]
  \centering
  \begin{threeparttable}
  \caption{Incentive structure by treatment}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c @{}}
  \toprule
  & \multicolumn{1}{c}{\textbf{Baseline}} & \multicolumn{1}{c}{\textbf{Bonus}} \\
  \midrule
  Referrer (sender) & Performance-based & Performance-based \\
  Referral (receiver) & No payment  & Fixed reward  \\
  \bottomrule
  \end{tabular*}
  \label{tab:treatments}
  \end{threeparttable}
\end{table}

We use a between-subjects design and randomly assign half our participants to the \textbf{Bonus} treatment. This allows us to causally identify the effect of the bonus on referral selection. Participants learn whether their referral gets the fixed bonus before making referral decisions.

\subsection{Belief elicitation}
We collect two sets of incentivized beliefs to assess the accuracy of participants' knowledge about exam performance. Participants first report beliefs about their own percentile ranking in the university for each exam area. After making referrals, participants report their beliefs about their referrals' percentile ranking in the university. For both belief elicitation tasks, participants earn \$5 per correct belief if their guess is within 7 percentiles of the true value. This margin of error is designed to balance precision with the difficulty of the task.

@@

\section{Sample, Incentives, and Procedure} \label{s4}
We invited all 4,417 undergraduate students who had completed their first year at the university at the time of recruitment to participate in our experiment. A total of 837 students participated in the data collection (19\% response rate). Our final sample consists of 734 individuals who referred peers with whom they had taken at least one class together, resulting in an 88\% success rate for the sample. We randomly allocated participants to either \textbf{Baseline} or \textbf{Bonus} treatments.

@@

The experiment was conducted online through Qualtrics, with participants recruited from active students. To ensure data quality while managing costs, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion. They also received potential earnings from one randomly selected belief question (up to \$5) and one randomly selected referral question (up to \$60). This structure resulted in maximum total earnings of \$82. The average time to complete the survey was 30 minutes, with an average compensation of \$80 for the one in ten participants randomly selected for payment. Payment processing occurred through bank transfer within 15 business days of participation.

@@

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Selection into the experiment}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
  & \multicolumn{1}{c}{\textbf{University}} & \multicolumn{1}{c}{\textbf{Sample}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\
  \midrule
  Reading score & 62.651 & 65.183 & $<0.001$ \\
  Math score & 63.973 & 67.477 & $<0.001$ \\
  GPA & 3.958 & 4.012 & $<0.001$ \\
  Low-SES & 0.343 & 0.410 & $<0.001$ \\
  Middle-SES & 0.505 & 0.499 & 0.763 \\
  High-SES & 0.153 & 0.091 & $<0.001$ \\
  Female & 0.567 & 0.530 & $<0.001$ \\
  Age & 21.154 & 20.651 & $<0.001$ \\
  \midrule
  Observations & 4,417 & 734 &  \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table compares characteristics between the university and the experimental sample. $p$-values for binary outcomes (Low-SES, Med-SES, High-SES, Female) are from two-sample tests of proportions; for continuous variables, from two-sample $t$-tests with unequal variances. All reported \textit{p}-values are two-tailed.
  \end{tablenotes}
  \label{tab:selection}
  \end{threeparttable}
\end{table}
